---
title: "Scoring single-cell data with CARTEx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring single-cell data with CARTEx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`cartex_sc()` applies the published CARTEx CD8+ T-cell exhaustion signature
(200 PCA-weighted genes) to **single-cell** data. It is designed to:

- read several formats — `.rds`, `.RData`, `.h5ad` (AnnData), `.h5Seurat`,
  `.loom` — or an in-memory Seurat object / matrix;
- run on a laptop: for the HDF5 formats it streams only the ~200 CARTEx genes
  off disk instead of loading the whole matrix or building a full object;
- be **transparent**: every decision (which layer was used, whether the data
  were raw counts or normalized, how genes were matched, what QC metadata is
  present) is captured in a `cartex_report` you can inspect.

## Installation

The HDF5 readers need the Bioconductor package **rhdf5** (it bundles its own
HDF5 library, so it installs cleanly on Mac/Windows/Linux):

```{r}
install.packages("BiocManager")
BiocManager::install("rhdf5")
# devtools::install_github("yourusername/CARTEx")
```

`.rds`/`.RData` Seurat objects only need `SeuratObject`.

## Quick start

```{r}
library(CARTEx)

# Any supported path, or an in-memory Seurat object / matrix
obj <- cartex_sc("pbmc.h5ad")

# Per-cell scores were added to the cell metadata:
head(obj$cartex_score)       # Z-scored across cells (headline value)
head(obj$cartex_score_raw)   # raw weighted dot-product

# The transparency report:
report <- attr(obj, "cartex_report")
print(report)
```

For a Seurat object you get the object back with `cartex_score` /
`cartex_score_raw` / `cartex_score_int` and a `cd8_status` column in `meta.data`,
ready for `FeaturePlot()`, `VlnPlot(obj, "cartex_score")`, etc. For the streaming
HDF5 path (no live Seurat object) you get a data.frame of scores joined to the
cell metadata.

## Restricting to CD8+ T cells

CARTEx is only valid for CD8+ T cells, so every cell is classified before
scoring. By default (`cd8 = "flag"`) all cells are scored and a `cd8_status`
column records the call; set `cd8 = "restrict"` to score only confident CD8+
T cells.

```{r}
# Score all cells but label them (default):
obj <- cartex_sc("pbmc.h5ad")
table(obj$cd8_status)

# Score ONLY confident CD8+ T cells:
cd8 <- cartex_sc("pbmc.h5ad", cd8 = "restrict")
```

How cells are classified (`cd8_method`):

- `"auto"` (default) — use a cell-type annotation column if one is recognizable
  (most reliable); otherwise gate on marker genes.
- `"metadata"` — annotation column only (set `cd8_column` to force a column).
- `"markers"` — gate on expression: CD8A/CD8B positive **and** CD3 positive
  **and** CD4 negative. (CD3 is required because NK cells also express CD8A.)
- `"none"` — skip classification.

Cells that cannot be resolved are labelled rather than guessed:
`ambiguous_dropout` (no CD8 or CD4 detected — common with sparse data),
`ambiguous_doublet` (both detected), `ambiguous_generic` (a bare "T cell"
label), or `non_T`. The report prints the full composition so you can see how
many cells were confidently CD8+. Marker gating is conservative: dropout means a
true CD8 cell can read zero for CD8A, which is exactly why the annotation path is
preferred when available.

## Pseudobulk analysis

When cluster/cell-type information is available you can score at the pseudobulk
level instead of per cell. Following the reference CARTEx pipeline, cells are
grouped, their **raw counts are summed**, the pseudobulk profiles are
LogNormalized, and then scored.

```{r}
# One pseudobulk profile per cluster:
pb <- cartex_sc("pbmc.h5ad", level = "pseudobulk", group_by = "seurat_clusters")

# CD8-only, with random replicate bins per group (as in the reference):
pb <- cartex_sc("pbmc.h5ad", level = "pseudobulk", group_by = "seurat_clusters",
                cd8 = "restrict", pseudo_replicates = 4)
```

`group_by` may name one or more metadata columns; if omitted, a cluster-like
column is auto-detected. With `pseudo_replicates > 1`, each group is randomly
split into that many replicate pseudobulk samples (reproducible via `seed`).
When the input is a Seurat object, each cell also receives its group's score in
`cartex_pseudobulk` for UMAP overlays, and the per-group table is attached as
`attr(obj, "cartex_pseudobulk_scores")`.

## How the score is defined

For each cell, the **raw** score is the dot product of its log-normalized
expression with the 200 CARTEx PCA weights. The headline `cartex_score` is that
value **Z-scored across all cells in the object** — matching the bulk
[`cartex_score()`] convention.

> **Caveat.** Because the Z-score is population-relative, `cartex_score` is only
> comparable *within* a single scored object, not across datasets scored
> separately. Use `cartex_score_raw` if you need an absolute per-cell quantity.

## Visualizing on a UMAP

`cartex_umap()` draws a side-by-side UMAP from a `cartex_sc()` result — the same
coordinates coloured by cell annotation (left) and by CARTEx score (right):

```{r}
res <- cartex_sc("pbmc.h5ad")
cartex_umap(res)                       # returns a ggplot/patchwork object
cartex_umap(res, file = "umap.png")    # or save to file
```

It reuses an existing UMAP embedding when one is present — a Seurat `umap`
reduction, an AnnData `obsm/X_umap`, or `UMAP_1`/`UMAP_2` metadata columns. If a
**Seurat object has no UMAP**, one is computed with default Seurat parameters
(`NormalizeData → FindVariableFeatures → ScaleData → RunPCA → RunUMAP`). (For the
streaming HDF5 path no full matrix is loaded, so a UMAP cannot be computed from
scratch — it must already exist in the file.)

## HTML report

`cartex_html_report()` writes a self-contained HTML summary of the run (no pandoc
required), optionally embedding the UMAP figure:

```{r}
cartex_html_report(res, file = "cartex_report.html")
```

The report captures input provenance, gene matching, CD8 composition, the
normalization decision, QC fields, the score summary, and any warnings.

## Transparency: checking your assumptions

The report exists because a per-cell score inherits every preprocessing choice
already baked into your object. It records:

- **Gene matching** — how many of the 200 genes were found and by which rule
  (exact symbol, cleaned symbol, or Ensembl ID), plus the unmatched list.
  CARTEx matches Ensembl IDs too, so CELLxGENE-style `.h5ad` files that key on
  Ensembl still score fully.
- **Normalization** — CARTEx inspects the actual values (not just the declared
  label, which is often wrong) to decide raw counts vs log-normalized, and only
  applies `LogNormalize` when the data look like counts. Library sizes for
  normalization are computed across *all* genes, even on the streaming path.
- **QC metadata** — detected `nFeature`/`nCount`/`percent.mt` columns with their
  ranges, and a heuristic flag for whether QC filtering looks already applied.
  If these are absent (as in many CELLxGENE exports), the report says so — a
  prompt to confirm upstream filtering yourself.

## Controlling behaviour

```{r}
# Force assume-normalized (skip detection):
cartex_sc("obj.rds", normalization = "assume_normalized")

# Force LogNormalize:
cartex_sc("obj.rds", normalization = "lognorm")

# Pick a specific assay / layer:
cartex_sc(seurat_obj, assay = "RNA", layer = "data")

# Just the report, or just the scores:
cartex_sc("obj.h5ad", return = "report")
cartex_sc("obj.h5ad", return = "scores")
```

## Bulk data

For bulk RNA-seq (genes × samples), keep using [`cartex_score()`] directly.
